### 2020-06-19 manuscript/main.tex
updated draft (2020-06-18)

\documentclass[letterpaper]{article}
\usepackage[letterpaper,top=1in,bottom=1in,left=1in,right=1in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage[T1]{fontenc}
\usepackage{times}      % Loads the Times-Roman Fonts
\usepackage{mathptmx}   % Loads the Times-Roman Math Fonts
\usepackage{helvet}
\usepackage[sf,bf]{titlesec}
\usepackage{lineno}
\usepackage{setspace}
\usepackage[style=nejm]{biblatex}
\addbibresource{references.bib}
\usepackage{caption}
\captionsetup[figure]{labelfont={sf,bf}}

@@

\textbf{Results:} The space of phylogenies inferred through \emph{profile sampling} reveals that consensus-inferred point estimates might not adequately represent them, both in topology and branch length. Cluster inference differed between Sanger- and NGS-derived sequences, and across gene regions; and only 30\% of clusters were detected unanimously.

@@

Public health officials and providers are interested in inferring transmission links between individuals with HIV to inform and improve HIV treatment and prevention \parencite{}. In the absence of reliable patient contact histories, phylogenetic analysis of HIV sequence data can and has been be used to infer transmission clusters \parencite{leitner}, under the assumption that two individuals sharing a most recent common ancestor in a phylogeny are more likely to share an epidemiological link in the real, unobservable transmission network. The application of phylogenetic analysis and cluster inference techniques in public health interventions to disrupt transmission was delineated as one of the four key pillars for achieving the Department of Health and Human Services' recently-announced plan for ending the HIV epidemic in the US \parencite{fauci}.

@@

HIV-1 \emph{pol} Sanger sequences (HXB2 positions 2253-3554), available through clinical care, were collected from the 37 adults ($\geq$18 years) newly-diagnosed with HIV during the first six months of 2013 and treated at The Miriam Hospital Immunology Center in Providence, Rhode Island, USA. Patients at this Center represent approximately 80% of the state's HIV epidemic.

In addition, blood specimens were obtained from consenting participants and processed to isolate peripheral blood mononuclear cells (PMBC) and plasma. Using both Sanger sequencing and NGS, near-whole genome viral sequences were obtained from one of those compartments; plasma for participants with detectable viral load and PBMC for participants with undetectable viral load or unsuccessful plasma genotyping. Total nucleic acids were extracted fand an in-house genotyping assay was used to generate the near-whole genome based on previously published methods \parencite{nadai, di_giallonardo}. For each sample, two cDNA templates were generated by SuperscriptIII First Strand Synthesis System (Thermofisher, Carlsbad, CA), followed by eight separate nested PCR reactions; these eight amplicons span the near-whole genome of HIV. Final amplicon products were sequenced by the Sanger method using 3100 Genetic Analyzer (Applied Biosystems, Foster City, CA) and by NGS using Nextera XT DNA Library Prep chemistry (Illumina, San Diego, CA) to generate multiplexed libraries for Illumina's MiSeq platform with 250 base paired-end reads. Sanger consensus sequences were generated manually using Sequencher version 5.2.4 (Gene Codes, Ann Arbor, MI) to confirm degenerate nucleotides. NGS data were processed and demultiplexed using the BaseSpace cloud application (Illumina, San Diego, CA).

@@

We introduce a new approach for incorporating within-host variation into phylogenetic analysis, called \emph{profile sampling}. We start by aligning each individual's NGS reads using the hivmmer pipeline \parencite{howison}, which we extended to support near-whole genome HIV data and to perform codon-aware alignment within each gene (pending release as hivmmer version 0.3.0). A key feature of this pipeline is its use of profile hidden Markov models (HMMs) to model and align collections of HIV sequences. Profile HMMs have been used in many kinds of biological sequence analyses and are particularly well-suited to modeling variation in populations of sequences \parencite{eddy2004}. Briefly, hivmmer performs quality control and error correction in overlapping regions of the read pairs using PEAR version 0.9.11 \parencite{zhang}, translates them into each possible reading frame, aligns them in amino acid space to profile HMMs of all group M reference sequences from the Los Alamos National Lab HIV Database \parencite{losalamos} using the profile HMM alignment tool HMMER version 3.1b2 \parencite{eddy2013}, and produces a codon frequency table across the near-whole HIV genome. We refer to this resulting codon frequency table as the individual's HIV \emph{profile}.

For each patient, we construct a population of fully-resolved sequences by sampling codons at each site in the genome using the codon frequencies from the \emph{profile}. The collection of sampled sequences captures the empirical distribution of within-host variation at the codon level. We then sample 500 sequences from each individual's \emph{profile}, and collate the sampled sequences into 500 \emph{profile-sampled} data sets, each having one sampled sequence per individual. We subsequently construct the 500 fully-resolved sequences per individual in order to use existing phylogenetic methods, because there is currently no published method to our knowledge for inferring a phylogeny directly from the profile representation of the individual's within-host variation.

@@

For profile sampling, we perform phylogenetic inference of near-whole genome sequences (wgs) on each of the 500 profile-sampled data sets by estimating a multiple sequence alignment with mafft version 7.313 \parencite{katoh} and a maximum-likelihood phylogeny with the GTRCAT model and 100 rapid bootstrap replicates using RAxML version 8.2.12 \parencite{stamatakis}. In addition to profile sampling, we infer two phylogenies with the same tools and parameters, one from the NGS consensus sequences and one from Sanger consensus sequences. We perform cluster inference on these phylogenies using Cluster Picker \parencite{ragonnet-cronin} with thresholds of 80\% bootstrap support and 4.5\% genetic distance. Analysis source code is available from \url{https://github.com/kantorlab/hiv-profile-sampling}.

Lastly, we also perform phylogenetic inference on three clinically relevant sub-genomic regions: the protease and reverse transcriptase regions at the beginning of the \emph{pol} gene (``prrt''), the \emph{int} gene, and the \emph{env} gene. The prrt and \emph{int} regions are routinely sequenced in clinical care to detect drug resistance mutations and inform clinical choices of anti-retroviral therapy. The \emph{env} region is sequenced to genotypically infer viral tropism and co-receptor usage.

We investigate the impact of within-host diversity on phylogenetic topology and evolutionary distance estimates. To examine the variation in topology, we first calculate the pairwise geodesic distance \parencite{billera,owen} among the 500 phylogenies from the profile samples, as well as the phylogenies from the NGS and Sanger consensus sequences. Then we perform multi-dimensional scaling (MDS) on the resulting distance matrix to visualize the topological space in two dimensions. Next, to examine the variation in estimated evolutionary distance, we sum the branch lengths within each phylogeny across all branches and across only the tip branches, and visualize the distribution of these branch length sums. Finally, we examine the clusters that are detected in phylogenies of NGS consensus sequences versus Sanger consensus sequences, and across the four genomic regions. We use the proportion of times a cluster appears across the 500 \emph{profile-sampled} phylogenies as a measure of support for that cluster. We refer to this value as the \emph{cluster support} and note that it is similar to the use of bootstrap support for evaluating the robustness of an individual phylogeny's topology, but extends that idea to evaluating the robustness of cluster detection using within-host sequence variation.

@@

Figure 1 shows the estimated within-host percent diversity in each examined genomic region across individuals, ordered by \emph{env}, which we expected \emph{a priori} to be the most variable region. The largest estimated diversity is in \emph{env} for individual MC28 (4.0\%), and \emph{env} has the overall largest range in estimated diversity (0.2\% to 4.0\%). The other regions have ranges of 0.2\% to 1.9\% (prrt), 0.1\% to 2.0\% (\emph{int}) and 0.2\% to 2.6\% (wgs). Such within-host estimations are not feasible via the conventional consensus Sanger or NGS approaches.

@@

Figure 2 demonstrates multi-dimensional scaling on the profile-sampled phylogenies and the phylogenies from the NGS and Sanger consensus sequences. The \emph{profile-sampling} approach reveals a multi-modal topological space in which the phylogenies inferred from the consensus sequences can be outliers. The prrt region displays four topological clusters, and the NGS and Sanger consensus phylogenies are located in different \emph{profile-sampled} clusters. There are two \emph{profile-sampled} clusters in \emph{int}, and the phylogeny of NGS consensus sequences lies at the center of one of the clusters, while the phylogeny of Sanger consensus sequences is an outlier. The phylogenies based on consensus sequences lie at opposite ends of the primary \emph{profile-sampled} MDS axis in \emph{env}, and of the secondary \emph{profile-sampled} MDS axis in the wgs region.

Figure 3 shows the distribution of branch length sums across the compared phylogenies. Overall, the estimates are larger in \emph{env} and the wgs regions, and smaller when restricting to only the tip branches. In some cases, the consensus phylogenies provide an adequate summary of the distribution (as in phylogeny of Sanger consensus sequences for all branches in \emph{env}, or the phylogeny of NGS consensus sequences for tip branches in \emph{env}). In other cases, the consensus phylogenies have estimates that are outliers in the distribution (as in both consensus phylogenies for either all or tip branches in the commonly-used prrt region).

Taken together, the heterogeneity between the phylogenetic results from profile sampling and the consensus-inferred point estimates demonstrate that within-host virus sequence diversity impacts the inference of virus phylogeny across patients, and that the consensus approach for collapsing within-host sequence variation can obscure both the magnitude and effect of these impacts. For example, the variation can result in multi-modal phylogenetic results and the point estimates from the phylogenies of consensus sequences can be outliers.

@@

The phylogeny of NGS consensus sequences detected eight clusters in prrt, nine in \emph{int}, four in \emph{env}, and six in wgs (Figure 4). Some clusters were robustly detected across all regions (e.g. the cluster (MC25, MC26, MC52), which had support between 98\% and 100\%), while other clusters were detected at lower \emph{cluster support}, and were not detected consistently across regions (e.g. cluster (MC14, MC59), which only appeared in the \emph{int} and wgs phylogenies, with \emph{cluster support} of 60\% to 62\%).

The phylogeny of Sanger consensus sequences detected eight clusters in prrt, seven in \emph{int}, six in \emph{env}, and seven in wgs (Figure 5). As with the phylogeny of NGS consensus sequences, the support and consistency of cluster detection varied across regions. Cluster detection was mostly consistent with the NGS consensus inference, except for the cluster (MC17, MC21), which only occurred in \emph{env} in the Sanger consensus, and had zero \emph{cluster support}. Three clusters were consistently detected across NGS and Sanger, and across all regions: (MC23, MC24); (MC25, MC26, MC52); and (MC27, MC28). Only 30\% of clusters were detected unanimously. Figure 6 provides a visual summary of the differences in cluster detection between Sanger- and NGS-derived sequences and across genomic regions.

@@

The diversity ranges we estimated are comparable to prior studies on within-host diversity \parencite{li, zanini}.

@@

Much of the enthusiasm about shifting from Sanger sequencing to NGS has been due to reducing costs and the ability to collect data on the entire HIV genome rather than a few genes. Our results suggest that much of the benefit of NGS may also be due to its greater depth of sequencing, which captures more variation in virus sequences within patients. This benefit can only be realized, though, if this variation is propagated to phylogenetic analyses, such as by the \emph{profile-sampling} method introduced here, rather than being collapsed to a consensus sequence, as is conventionally done.

@@

In our comparison of cluster inference across genomic regions, we found that fewer clusters were detected overall in \emph{env} and wgs compared to prrt and \emph{int}. Prior studies of clustering from Sanger consensus sequences present mixed results on the prevalence of clustering across genomic regions. Some studies have found concordant clustering across \emph{gag}-\emph{env} \parencite{han} and \emph{gag}-\emph{pol}-\emph{env} \parencite{english,kaye}, while others found fewer clusters in \emph{pol} than in \emph{env} \parencite{kapaata}, or fewer clusters in \emph{gag}-\emph{env} than in \emph{pol} \parencite{ndiaye}. The additional information available in deep-sequenced NGS data, along with the cluster support measures provided by profile sampling, could help resolve these differences, as suggested here. In our data, not only were more clusters detected in the \emph{pol} region, but those clusters had higher cluster support as measured by deep-sequenced NGS data.

One limitation of our study is the small number of participants and the short time-frame patients were enrolled in. The participants enrolled here represent a dense temporal sampling, and comprise all newly HIV-diagnosed individuals in a six month period at the largest HIV center in Rhode Island, USA, in which 80\% of the state's people with HIV are cared for. The overall size of the HIV epidemic in Rhode Island was estimated as 2,396 individuals in 2016 \parencite{ridoh}, but NGS data for this population are not currently available beyond those presented here. In future work, we will apply \emph{profile sampling} to larger NGS data sets, to assess cluster inference concordance between Sanger and NGS data, especially for clusters larger than three individuals, which was the largest cluster size here.

@@

This work was supported by the National Institutes of Health [R01AI136058, K24AI134359, P30AI042853, P20GM109035 ] and a Brown University DEANS Award.

@@

\begin{figure}[p!]
	\caption{Intra-patient genetic diversity (defined as the average percent difference across all pairwise comparisons of the 500 profile-sampled nucleotide sequences for a patient) is highest in \emph{env} for most individuals, and lies within the range of previously reported values.}
	\centering
	\includegraphics[width=\linewidth]{Figure1}
	\label{fig1}
\end{figure}

\begin{figure}[p!]
	\caption{Multi-dimensional scaling of pairwise geodesic distance among maximum-likelihood phylogenies from the profile-sampling approach show that the space of phylogenies inferred for the prrt and int regions are multi-modal. The phylogenies from consensus sequences (black dots) are point estimates that do not capture the full variation in phylogenies that can be inferred from deeply-sequenced NGS data.}
	\centering
	\includegraphics[width=\linewidth]{Figure2}
	\label{fig2}
\end{figure}

\begin{figure}[p!]
	\caption{The total branch length in each of the profile-sampled phylogenies also varies. The phylogenies from consensus sequences (black dots) can lie at extreme values within these distribution, both when considering the lengths across all branches and the lengths across only the branches at the tips.}
	\centering
	\includegraphics[width=3.25in]{Figure3}
	\label{fig3}
\end{figure}

\begin{figure}[p!]
	\caption{Clusters (vertical colored bars) inferred from the phylogenies of NGS consensus sequences differ across genomic regions. The largest number of clusters was inferred from the int region, and the smallest number from the env region. Profile sampling provides a bootstrapped measure of cluster support (annotation to bars).}
	\centering
	\includegraphics[width=\linewidth]{Figure4}
	\label{fig4}
\end{figure}

\begin{figure}[p!]
	\caption{Clusters (vertical colored bars) inferred from the phylogenies of Sanger consensus sequences differ across genomic regions. The largest number of clusters was inferred from the int region, and the smallest number from the env region. Profile sampling provides a bootstrapped measure of cluster support (annotation to bars).}
	\centering
	\includegraphics[width=\linewidth]{Figure5}
	\label{fig5}
\end{figure}

\newpage

\begin{figure}[p!]
	\caption{Summary of clusters identified by Sanger versus NGS consensus sequences across genomic regions. Numeric values indicate bootstrapped cluster support from the profile sampling method. A blank cell indicates that the cluster was not detected in that consensus sequence and genomic region.}
	\centering
	\includegraphics[width=3.25in]{Figure6}
	\label{fig6}
\end{figure}

>>>>

\documentclass[letterpaper]{article}
\usepackage[letterpaper,top=1in,bottom=1in,left=1in,right=1in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage[T1]{fontenc}
\usepackage{times}      % Loads the Times-Roman Fonts
\usepackage{mathptmx}   % Loads the Times-Roman Math Fonts
\usepackage{helvet}
\usepackage[sf,bf]{titlesec}
\usepackage{lineno}
\usepackage{setspace}
\usepackage[style=nejm]{biblatex}
\addbibresource{references.bib}
\usepackage{caption}
\usepackage{float}
\captionsetup[figure]{labelfont={sf,bf}}

@@

\textbf{Results:} The space of phylogenies inferred through \emph{profile sampling} reveals that consensus-inferred point estimates might not adequately represent them, both in topology and branch length. Cluster inference differed between Sanger- and NGS-derived sequences, and across gene regions. Only one of twelve of clusters was detected unanimously. Cluster support, however, was unanimously better when using the near-whole genome.

@@

Public health officials and providers are interested in inferring transmission links between individuals with HIV to inform and improve HIV treatment and prevention \parencite{hogben}. In the absence of reliable patient contact histories, phylogenetic analysis of HIV sequence data can and has been be used to infer transmission clusters \parencite{leitner}, under the assumption that two individuals sharing a most recent common ancestor in a phylogeny are more likely to share an epidemiological link in the real, unobservable transmission network. The application of phylogenetic analysis and cluster inference techniques in public health interventions to disrupt transmission was delineated as one of the four key pillars for achieving the Department of Health and Human Services' recently-announced plan for ending the HIV epidemic in the US \parencite{fauci}.

@@

HIV-1 \emph{pol} Sanger sequences (HXB2 positions 2253-3554), available through clinical care, were collected from the 37 adults ($\geq$18 years) newly-diagnosed with HIV during the first six months of 2013 and treated at The Miriam Hospital Immunology Center in Providence, Rhode Island, USA. Patients at this Center represent approximately 80\% of the state's HIV epidemic.

In addition, blood specimens were obtained from consenting participants and processed to isolate peripheral blood mononuclear cells (PMBC) and plasma. Using both Sanger sequencing and NGS, near-whole genome viral sequences were obtained from one of those compartments; plasma for participants with detectable viral load and PBMC for participants with undetectable viral load or unsuccessful plasma genotyping. Total nucleic acids were extracted fand an in-house genotyping assay was used to generate the near-whole genome based on previously published methods \parencite{nadai, di_giallonardo}. For each sample, two cDNA templates were generated by SuperscriptIII First Strand Synthesis System (Thermofisher, Carlsbad, CA), followed by eight separate nested PCR reactions; these eight amplicons span the near-whole genome of HIV. Final amplicon products were sequenced by the Sanger method using 3100 Genetic Analyzer (Applied Biosystems, Foster City, CA) and by NGS using Nextera XT DNA Library Prep chemistry (Illumina, San Diego, CA) to generate multiplexed libraries for Illumina's MiSeq platform with 250 base paired-end reads. Sanger consensus sequences were generated manually using Sequencher version 5.2.4 (Gene Codes, Ann Arbor, MI) to confirm degenerate nucleotides. NGS data were processed and demultiplexed using the BaseSpace cloud application (Illumina, San Diego, CA). Both Sanger and NGS consensus sequences were called at a 20\% threshold, meaning that any site with two or more nucleotides with $\geq 20\%$ frequency is represented by an ambiguous nucleotide code.

@@

We introduce a new approach for incorporating within-host variation into phylogenetic analysis, called \emph{profile sampling}. We start by aligning each individual's NGS reads using the hivmmer pipeline \parencite{howison}, which we extended to support near-whole genome HIV data and to perform codon-aware alignment within each gene (hivmmer version 0.2.1). A key feature of this pipeline is its use of profile hidden Markov models (HMMs) to model and align collections of HIV sequences. Profile HMMs have been used in many kinds of biological sequence analyses and are particularly well-suited to modeling variation in populations of sequences \parencite{eddy2004}. Briefly, hivmmer performs quality control and error correction in overlapping regions of the read pairs using PEAR version 0.9.11 \parencite{zhang}, translates them into each possible reading frame, aligns them in amino acid space to profile HMMs of all group M reference sequences from the Los Alamos National Lab HIV Database \parencite{losalamos} using the profile HMM alignment tool HMMER version 3.1b2 \parencite{eddy2013}, and produces a codon frequency table across the near-whole HIV genome. We refer to this resulting codon frequency table as the individual's HIV \emph{profile}.

For each individual, we construct a population of fully-resolved sequences by sampling codons at each site in the genome using the codon frequencies from the \emph{profile}. The collection of sampled sequences captures the empirical distribution of within-host variation at the codon level. We then sample 500 sequences from each individual's \emph{profile}, and collate the sampled sequences into 500 \emph{profile-sampled} data sets, each having one sampled sequence per individual. We subsequently construct the 500 fully-resolved sequences per individual in order to use existing phylogenetic methods, because there is currently no published method to our knowledge for inferring a phylogeny directly from the profile representation of the individual's within-host variation.

@@

For profile sampling, we perform phylogenetic inference of near-whole genome sequences (wgs) on each of the 500 profile-sampled data sets by estimating a multiple sequence alignment with OMM\_MACSE version 10.02 \parencite{ranwez} and a maximum-likelihood phylogeny with the GTRCAT model and 100 rapid bootstrap replicates using RAxML version 8.2.12 \parencite{stamatakis}, with an HIV group O sequence (GenBank accession L20587.1) as the outgroup. We also perform this same phylogenetic inference on three clinically relevant sub-genomic regions: the protease and reverse transcriptase regions at the beginning of the \emph{pol} gene (``prrt''), the \emph{int} gene, and the \emph{env} gene. The prrt and \emph{int} regions are routinely sequenced in clinical care to detect drug resistance mutations and inform clinical choices of anti-retroviral therapy. The \emph{env} region is sequenced to genotypically infer viral tropism and co-receptor usage.

In addition to profile sampling, we infer phylogenies with the same tools, regions and parameters for the NGS consensus sequences and the Sanger consensus sequences. We perform cluster inference on all phylogenies using Cluster Picker \parencite{ragonnet-cronin} with a threshold of 99\% bootstrap support. We do not impose a genetic distance threshold because empirically-justified thresholds that are comparable across the near-whole genome and the prrt, \emph{int}, and \emph{env} regions have not to our knowledge been established. This approach of using only a bootstrap criteria for cluster detection is consistent with methods commonly used in the broader literature on HIV cluster analysis \parencite{hassan}.

We investigate the impact of within-host diversity on phylogenetic topology and evolutionary distance estimates. To examine the variation in topology, we first calculate the pairwise geodesic distance \parencite{billera,owen} among the 500 phylogenies from the profile samples, as well as the phylogenies from the NGS and Sanger consensus sequences. Then we perform multi-dimensional scaling (MDS) on the resulting distance matrix to visualize the topological space in two dimensions. Next, to examine the variation in estimated evolutionary distance, we sum the branch lengths within each phylogeny across all branches and across only the tip branches, and visualize the distribution of these branch length sums. Finally, we examine the clusters that are detected in phylogenies of NGS consensus sequences versus Sanger consensus sequences, and across the four genomic regions. We use the frequency that a cluster appears across the 500 \emph{profile-sampled} phylogenies as a measure of support for that cluster. We refer to this value as the \emph{cluster support} and note that it is similar to the use of bootstrap support for evaluating the robustness of an individual phylogeny's topology, but extends that idea to evaluating the robustness of cluster detection using within-host sequence variation.

All analysis source code is available from \url{https://github.com/kantorlab/hiv-profile-sampling}.

@@

Figure 1 shows the estimated within-host percent diversity in each examined genomic region across individuals, ordered by \emph{env}, which we expected \emph{a priori} to be the most variable region. The largest estimated diversity is in \emph{env} for individual MC28 (3.9\%), and \emph{env} has the overall largest range in estimated diversity (0.2\% to 3.9\%). The other regions have ranges of 0.2\% to 1.9\% (prrt), 0.1\% to 2.0\% (\emph{int}) and 0.2\% to 2.6\% (wgs). Such within-host estimations are not feasible via the conventional consensus Sanger or NGS approaches.

@@

Figure 2 demonstrates multi-dimensional scaling on the \emph{profile-sampled} phylogenies and the phylogenies from the NGS and Sanger consensus sequences within each genomic region (see Figure S1 for multi-dimensional scaling across all regions). The \emph{profile-sampling} approach reveals for each of the genomic regions a multi-modal topological space in which the phylogenies inferred from both the NGS and Sanger consensus sequences are outliers. A key different between the consensus and \emph{profile-sampled} sequences is that the consensus sequences contain ambiguous nucleotide codes at sites with two or more nucleotides with $\geq 20\%$ frequency. In contrast, \emph{profile-sampled} sequences by construction have no ambiguous sites, and ambiguity is instead incorporated into the analysis through the frequency of the ambiguous nucelotides across the 500 samples.

Figure 3 shows the distribution of branch length sums across the compared phylogenies. Overall, the estimates are larger in \emph{env} and the wgs regions, and smaller when restricting to only the tip branches. In some cases, the consensus phylogenies provide an adequate summary of the distribution (as in the phylogeny of the NGS consensus sequence for tip branches for wgs). In other cases, the consensus phylogenies have estimates that are outliers in the distribution (as in the phylogenies from NGS and Sanger consensus sequences for all branches in wgs and \emph{env}).

Taken together, the heterogeneity between the phylogenetic results from profile sampling and the consensus-inferred point estimates demonstrate that within-host virus sequence diversity impacts the inference of virus phylogeny across individuals, and that the consensus approach to handling ambiguity and collapsing within-host sequence variation can obscure both the magnitude and effect of these impacts.

@@

Figure 4 illustrates the difference in cluster support across sequencing methods and genomic regions. Some clusters have consistently high support across all methods and regions, such as MC25/MC26/MC52 and MC14/MC59. Other clusters had higher support in certain genomic regions, such as MC17/MC20/MC21 in \emph{env} and wgs. One larger cluster, MC37/MC41/MC47/MC53/MC56, appeared only in the wgs region, and only about half of the time. Furthermore, the clusters detected in the wgs region had the highest overall clsuter support for every cluster.

Figure 5 compares the differences in cluster detection between the phylogenies from the Sanger and the NGS consensus sequences. The phylogenies of NGS consensus sequences detected six clusters in prrt, seven in \emph{int}, eight in \emph{env}, and nine in wgs (Figure S2). Five clusters were robustly detected across all regions, while other clusters were detected at lower \emph{cluster support}, and were not detected consistently across regions (e.g. cluster (MC14, MC59), which only appeared in the \emph{int} and wgs phylogenies, with \emph{cluster support} of 60\% to 62\%).

The phylogenies of Sanger consensus sequences detected fewer clusters than the phylogenies of NGS consensus sequence in all regions except \emph{env}. The phylogenies of Sanger consensus sequences detected four clusters in prrt, six in \emph{int}, nine in \emph{env}, and eight in wgs (Figure S3). Only one cluster (MC25/MC26/MC52) was consistently detected across the phylogenies from NGS and Sanger consensus sequences, and across all regions. Neither detected the larger cluster, MC37/MC41/MC47/MC53/MC56, which was detected only through profile sampling.

@@

@@

Much of the enthusiasm about shifting from Sanger sequencing to NGS has been due to reducing costs and the ability to collect data on the entire HIV genome rather than a few genes. Our results suggest that much of the benefit of NGS may also be due to its greater depth of sequencing, which captures more variation in virus sequences within individuals. This benefit can only be realized, though, if this variation is propagated to phylogenetic analyses, such as by the \emph{profile-sampling} method introduced here, rather than being collapsed to a consensus sequence, as is conventionally done.

@@

In our comparison of cluster inference across genomic regions, we found that fewer clusters were detected overall in prrt and \emph{int} compared to \emph{env} and wgs. Prior studies of clustering from Sanger consensus sequences present mixed results on the prevalence of clustering across genomic regions. Some studies have found concordant clustering across \emph{gag}-\emph{env} \parencite{han} and \emph{gag}-\emph{pol}-\emph{env} \parencite{english,kaye}, while others found fewer clusters in \emph{pol} than in \emph{env} \parencite{kapaata}, or fewer clusters in \emph{gag}-\emph{env} than in \emph{pol} \parencite{ndiaye}. The additional information available in deep-sequenced NGS data, along with the cluster support measures provided by profile sampling, could help resolve these differences, as suggested here. In our data, not only were more clusters detected in the near-whole length genome, but those clusters also had higher cluster support as measured by deep-sequenced NGS data.

One limitation of our study is the small number of participants and the short timeframe they were enrolled in. The participants enrolled here represent a dense temporal sampling, and comprise all newly HIV-diagnosed individuals in a six month period at the largest HIV center in Rhode Island, USA, in which 80\% of the state's people with HIV are cared for. The overall size of the HIV epidemic in Rhode Island was estimated as 2,396 individuals in 2016 \parencite{ridoh}, but NGS data for this population are not currently available beyond those presented here. In future work, we will apply \emph{profile sampling} to larger NGS data sets, to assess cluster inference concordance between Sanger and NGS data, especially for clusters larger than five individuals, which was the largest cluster size here.

@@

This work was supported by the National Institutes of Health [R01AI136058, K24AI134359, P30AI042853, P20GM109035] and a Brown University DEANS Award.

@@

\begin{figure}[H]
	\caption{Intra-host genetic diversity (defined as the average percent difference across all pairwise comparisons of the 500 profile-sampled nucleotide sequences for an individual) is highest in \emph{env} for most individuals, and lies within the range of previously reported values.}
	\centering
	\includegraphics[width=6.5in]{Figure1.emf}
\end{figure}

\begin{figure}[H]
	\caption{Multi-dimensional scaling of pairwise geodesic distance among maximum-likelihood phylogenies from the profile-sampling approach show that the space of inferred phylogenies is multi-modal for all genomic regions. The phylogenies from consensus sequences (dot and triangle) are point estimates that do not capture the full variation in phylogenies that can be inferred from deeply-sequenced NGS data.}
	\centering
	\includegraphics[width=6.5in]{Figure2.emf}
\end{figure}

\begin{figure}[H]
	\caption{The total branch length in each of the profile-sampled phylogenies also varies. The phylogenies from consensus sequences (dot and triangle) can lie at extreme values within these distribution, both when considering the lengths across all branches and the lengths across only the branches at the tips.}
	\centering
	\includegraphics[width=6.5in]{Figure3.emf}
\end{figure}

\begin{figure}[H]
	\caption{The cluster support of clusters identified by profile sampling across genomic regions. A blank cell indicates that the cluster was not detected in that genomic region.}
	\centering
	\includegraphics[width=6.5in]{Figure4.emf}
\end{figure}

\begin{figure}[H]
	\caption{Summary of clusters identified by Sanger versus NGS consensus sequences across genomic regions. Numeric values indicate cluster support from the profile sampling method. A blank cell indicates that the cluster was not detected in that genomic region.}
	\centering
	\includegraphics[width=6.5in]{Figure5.emf}
\end{figure}

\begin{figure}[H]
	\caption{Multi-dimensional scaling of pairwise geodesic distance among maximum-likelihood phylogenies from the profile-sampling approach shows that the space of inferred phylogenies are multi-modal for all genomic regions. The phylogenies from consensus sequences are point estimates that do not capture the full variation in phylogenies that can be inferred from deeply-sequenced NGS data.}
	\centering
	\includegraphics[width=6.5in]{FigureS1.emf}
\end{figure}

\begin{figure}[H]
	\caption{Clusters (vertical colored bars) inferred from the phylogenies of NGS consensus sequences differ across genomic regions. The largest number of clusters was inferred from the wgs region, and the smallest number from the prrt region. Profile sampling provides a bootstrapped measure of cluster support (annotation to bars).}
	\centering
	\includegraphics[width=6.5in]{FigureS2.emf}
\end{figure}

\begin{figure}[H]
	\caption{Clusters (vertical colored bars) inferred from the phylogenies of Sanger consensus sequences differ across genomic regions. The largest number of clusters was inferred from the env region, and the smallest number from the prrt region. Profile sampling provides a bootstrapped measure of cluster support (annotation to bars).}
	\centering
	\includegraphics[width=6.5in]{FigureS3.emf}
\end{figure}
